Verification of Stencil Computations through Spatial Specifications

### 2016-07-03 paper.tex
started to count bounded/exact

\definecolor{darkblue}{rgb}{0.0,0.0,0.5}
\definecolor{darkgreen}{rgb}{0.0,0.4,0.0}
\definecolor{darkdarkgreen}{rgb}{0.0,0.35,0.0}

\newcommand{\sten}[1]{\textcolor{darkdarkgreen}{\!#1}}

@@

\CustomVerbatimEnvironment{SpecVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,
xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98,numbersep=0.9em}
\CustomVerbatimEnvironment{ExmVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,
xrightmargin=0.2cm,baselinestretch=0.98,numbers=left,numbersep=0.9em,commandchars=\\\{\}}
\CustomVerbatimEnvironment{IVerbatim}{Verbatim}{fontsize=\relsize{-1},xleftmargin=0.5cm,
xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98,numbersep=0.9em}

@@

\newcommand{\dnote}[1]{\textcolor{darkpurple}{Dom: #1}}
\newcommand{\mnote}[1]{\textcolor{darkgreen}{Mistral: #1}}

@@

x = a(i, j)   + a(i-1, j)   + a(i+1, j)
y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
b(i, j) = (x + y + z) / 9.0
\end{ExmVerbatim}
%%
Such a stencil computation is a common image convolution (for example
for edge detection). 
Note that there are a number of intermediate assignments over which
the whole specification ranges.

As another example, the following code and specification
implements the \emph{Roberts cross}
edge-detection convolution~\cite{davis1975survey}:
%%
\begin{ExmVerbatim}
do j=0, jmax-1
    do i=0, imax-1
\sten{      != region r1 = forward(depth=1, dim=1)}
\sten{      != region r2 = forward(depth=1, dim=2)}
\sten{      != stencil readOnce, r1*r2 :: a}
      x = a(i,j) - a(i+1,j+1)
      y = a(i+1,j) - a(i,j+1)
    b(i,j) = sqrt(x*x+y*y)
   end do
end do
\end{ExmVerbatim}

@@

\begin{figure}
\begin{tabular}{|l|l|l|}
\hline
Package       & \% potential & \% actual   \\
\hline
UM            & \umtickAssignPercent &  \umtickAssignSuccessPercent   \\
E3MG          & \ethreemgeatickAssignPercent & \ethreemgeatickAssignSuccessPercent\\
BLAS          & \blastickAssignPercent & \blastickAssignSuccessPercent \\
Hybrid4       & \hybridfourtickAssignPercent & \hybridfourtickAssignPercent \\
GEOS-Chem     & \geoschemtickAssignPercent & \geoschemtickAssignPercent \\
Navier        & \naviertickAssignPercent & \naviertickAssignPercent \\
CP            & \computationalphysicstwotickAssignPercent & \computationalphysicstwotickAssignSuccessPercent\\
\hline
\textbf{Total}& \overalltickAssignPercent  & \overalltickAssignSuccessPercent    \\
\hline 
\end{tabular}
\caption{Stencil assignment rates\label{fig:tickAssign}}
\end{figure}

@@

\subsection{Related work}

Various prior approaches to specifying and verifying the behaviour
of stencils have been based on fine-grained representations of the
code. These essentially duplicate the indexing scheme of a stencil
computation in their specifications.

\citet{kamil2016verified} propose a technique they call
\emph{verified lifting} for mapping low-level, possibly
greatly optimised, stencil computations into a high-level
predicate language using inductive program synthesis. 
The aim is that these high-level abstract specifications can
then be translated into high-performance DSLs, easing the
use of high-performance DSLs. Compared to our approach,
their specification language is more fine grained, capturing
the exact indexing pattern of a stencil. For example, the following
is an example post-condition in their language, generated by their
compiler \textsc{stng} from a Fortran stencil computation~\cite [p.3]{kamil2016verified}:
%
\begin{align*}
& \textit{post}(\texttt{a}, \texttt{b}) \equiv \forall \texttt{imin}+1 
\leq \texttt{i} \leq \texttt{imax}, \texttt{jmin} \leq \texttt{j} \leq
\texttt{jmax} \\
& \qquad \texttt{a}(\texttt{i},\texttt{j}) = 
\texttt{b}(\texttt{i}-1,\texttt{j}) + \texttt{b}(\texttt{i},\texttt{j})
\end{align*}
%
Noticeable, the bounds of the stencil and the exact indexing pattern
are captured. Our approach aims to be much more abstract, to
facilitate human reasoning, and avoid low-level lexical
errors. However, \textsf{stng} specifications are generated, \ie{}, 
are never written by a programmer, avoiding this
problem. Their inductive program synthesis approach allows
specifications to extracted even from optimised and unrolled code. For
example, the above post condition is generated from the code~\cite [p.3]{kamil2016verified}:
%
\begin{ExmVerbatim}
do j=jmin, jmax 
    t = b(imin, j)
    do i=imin+1,imax
        q = b(i, j)
        a(i, j) = q + t
        t = q
    end do
end do
\end{ExmVerbatim}
%
Thus, \textsc{stng} can understand cross-loop data-dependencies
in order to assign the above post-condition. CamFort is not currently
able to ascribe a specification in the above case as the use of
\texttt{t} on line 5 maps only to the definition on line 2, and no on
line 6. Extending our approach to follow loop-carried dependencies is
future work. An alternate approach for us would be to build ontop of
the work of Kamil~\emph{et al.} by inferring our specifications from 
\textsc{stng} (as an intermediate step).

In an alternate approach,~\citet{abe2013model}

There are many DSLs for stencil comptuation aimed at either
performance or verification, or both. Many elimiante out-of-bounds
errors, \eg{},~\cite{DBLP:journals/corr/abs-1109-0777}.

>>>>

\definecolor{darkblue}{rgb}{0.0,0.0,0.5}
\definecolor{darkgreen}{rgb}{0.0,0.4,0.0}

@@

\CustomVerbatimEnvironment{SpecVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,
xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98,numbersep=0.9em}
\CustomVerbatimEnvironment{ExmVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,
xrightmargin=0.2cm,baselinestretch=0.98,numbers=left,numbersep=0.9em}
\CustomVerbatimEnvironment{IVerbatim}{Verbatim}{fontsize=\relsize{-1},xleftmargin=0.5cm,
xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98,numbersep=0.9em}

@@

\newcommand{\dnote}[1]{\textcolor{darkpurple}{Dom: #1}}
\newcommand{\mnote}[1]{\textcolor{darkgreen}{Mistral: #1}}
\newcommand{\anote}[1]{\textcolor{red}{Andy: #1}}

@@

x = a(i, j)   + a(i-1, j)   + a(i+1, j)
y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
b(i, j) = (x + y + z) / 9.0
\end{ExmVerbatim}
%%
Note that there are a number of intermediate assignments over which
the whole specification ranges.

@@

\begin{figure}
\begin{tabular}{|l|l|l|l|l|}
\hline
Package       & \% potential & \% actual & Total & Total    \\
              & stencils     & stencils  & exact  & bounded \\
\hline
UM            & \umtickAssignPercent & \umtickAssignSuccessPercent & \umnumStencilLines & \umboundedBoth  \\
E3MG          & \ethreemgeatickAssignPercent & \ethreemgeatickAssignSuccessPercent & \ethreemgeanumStencilLines & \ethreemgeaboundedBoth \\
BLAS          & \blastickAssignPercent & \blastickAssignSuccessPercent & \blasnumStencilLines & \blasboundedBoth \\
Hybrid4       & \hybridfourtickAssignPercent & \hybridfourtickAssignPercent & \hybridfournumStencilLines &  \hybridfourboundedBoth\\
GEOS-Chem     & \geoschemtickAssignPercent & \geoschemtickAssignPercent & \geoschemnumStencilLines & \geoschemboundedBoth \\
Navier        & \naviertickAssignPercent & \naviertickAssignPercent & \naviernumStencilLines & \navierboundedBoth \\
CP            & \computationalphysicstwotickAssignPercent & \computationalphysicstwotickAssignSuccessPercent & \computationalphysicstwonumStencilLines & \computationalphysicstwoboundedBoth \\
\hline
\textbf{Total}& \overalltickAssignPercent  & \overalltickAssignSuccessPercent & \overallnumStencilLines & \overallboundedBoth    \\
\hline 
\end{tabular}
\caption{Stencil assignment rates\label{fig:tickAssign}}
\end{figure}

The first two columns of Figure \ref{fig:tickAssign} shows the percentage of lines which were
assigned potential and actual stencils.
\anote{Add explanation about why our results are low. Include an example from BLAS of a tickAssign'd statement that is not a stencil}
\anote{Add discussion about the rate of stencils in the code. Comparison with rate of assignment statements would be nice}

If the inference process cannot determine an exact stencil for a statement it will instead produce an upper (\term{atMost}) and lower (\term{atLeast}) bound. The numbers of exact and bounded stencils are shown in columns three and four of Figure \ref{fig:tickAssign}. We can see that our stencil language is capable of precisely specifying the overwhelming majority of stencils. Most of the bounded stencils apply to some form of data copying. For example (taken from E3MG):
\begin{ExmVerbatim}
!- stencil atLeast, readOnce, irreflexive(dims=2), (reflexive(dim=1)) :: sfda
!- stencil atMost, readOnce, (forward(depth=8, dim=2))*(reflexive(dim=1)) :: sfdt
ZZ1(I,J) = SFDT(I,J+8)
\end{ExmVerbatim}

@@
